Source-dependent likelihood-ratio ranking statistic for compact binary coalescence candidates

### 2016-10-07 article_rankstatsrc.tex
Added new plots with BBH

The direct detections of gravitational waves by the Advanced Laser Interferometer Gravitational-wave Observatory (Advanced LIGO) that were made in late 2015 \cite{Abbott:2016blz, Abbott:2016nmj, TheLIGOScientific:2016pea} mark a monumental breakthrough in physics and astronomy. With additional Advanced LIGO observing runs scheduled and further gravitational-wave observatories expected to come online in in the next several years, the age of gravitational-wave astronomy has truly begun. Advanced LIGO's design sensitivity aims to achieve a factor of three improvement to the signal-to-noise ratio (SNR) for binary black hole systems similar to the detected events \cite{TheLIGOScientific:2016agk}. These improvements are expected to increase the inferred rate of observed compact binary coalescences \cite{Abbott:2016nhf}. With additional detections occurring more and more frequently, we stand to gain a deeper understanding of these tumultuous and dramatic astrophysical events.

@@

In Section \ref{sec:rankingstat}, we begin by describing the ranking statistic that is currently used in Advanced LIGO's low-latency searches for compact binary coalescences. The derivation of a ranking statistic that envelops astrophysical source model information is outlined in Section \ref{sec:derivation}, where the various terms are explored in further detail in Sections \ref{sec:P_j} and \ref{sec:P_k}. In Section \ref{sec:P_jk}, we present a discussion of the source-dependent ranking statistic and test it using a BNS and BBH source population model. We conclude in Section \ref{sec:conclusion}.

@@

We consider a gravitational-wave candidate $\vec{s}$; it belongs to an astrophysical source population $\mathcal{S}$ and has parameters $\vec\theta$:
   \begin{equation}
   \vec\theta = \{m_1,m_2,\chi_1,\chi_2,\rho\},
   \label{eqn:theta}
   \end{equation}
where $m_i$ and $\chi_i$ are the masses and spins, respectively, of source $i=1,2$ and $\rho$ is the nominal SNR.

@@

The first term in the summation of Eq. \ref{eqn:prob_signal} depends on the astrophysical source population model; therefore, a choice must be made for the source population model. In this work, we will consider two models: a binary neutron star (BNS) population and a binary black hole (BBH) population. The BNS source population we used is from Ozel {\it et al.} \cite{Ozel:2012ax}, where neutron star masses are centred around 1.33+1.33$M_\odot$ with a dispersion of 0.05$M_\odot$. For the BBH source population, we take a distribution that is centred around 30+30$M_{\odot}$ with a dispersion of 5$M_\odot$. The population models do not consider the spins of the sources.

The probability density function (PDF) of the source population is given as $P(\vec{\gamma})$, where $\vec{\gamma}=\{\gamma_1,\gamma_2,..., \gamma_M\}$ are the intrinsic parameters of the source population, such as the masses and spins. The probability of that a signal looks like $\vec{t}_j$ (multiplied by some $\rho$) is the following:
	\begin{align}
	P(\vec{t}_j|\text{signal}) &= \int P_j(\vec{\gamma}) \diff\vec{\gamma} \\
	&\approx P_j(\gamma_1) P_j(\gamma_2)...P_j(\gamma_M) \times V_j.
	\end{align}
In the final equality, we assume that the distribution of the source parameters are independent of one another, so the joint probability is the product of the marginals. Because the template bank is made of discrete waveforms, the template bank is turned into a Voronoi diagram, where each template is represented by a Voronoi region.$V_j$ is the $M$-dimensional volume of the $j$th region of the Voronoi diagram, computed using Delaunay triangulation, and it is assumed that $V_j$ is small enough such that $P_j(\gamma_i)$ is constant within the patch. We compute $P(\vec{t}_j|\text{signal})$ for each template in the template bank to find the probability that a system from the given source population will have the same properties as the template. Results for both source populations are shown in Figure \ref{fig:p_j}.

@@

@@

We assume that $P(\rho)\propto\rho^{-4}$. We cannot normalize this because there would be an infinite density of 0 SNR sources at infinite distance. Instead, we normalize $\rho^{-4}$, from $\rho_{\text{low}}=4$, the trigger threshold for a gravitational-wave search, to $\rho_{\text{high}}=\infty$.

\begin{figure}
\includegraphics[width=0.5\textwidth]{logP_Ozelrho0_0.png}
\includegraphics[width=0.5\textwidth]{logP_Ozelrho16_3802850078.png}
\includegraphics[width=0.5\textwidth]{logP_Ozelrho51_2132034356.png}
\caption{log($P(\vec{t}_k|\text{signal},\rho)$) for different values of $\rho$ for the BNS source population. The solid black lines denote the mass distribution of the respective source population model.}
\label{fig:P_tkj_BNS}
\end{figure}
\begin{figure}
\includegraphics[width=0.5\textwidth]{logP_Kiziltanrho0_0.png}
\includegraphics[width=0.5\textwidth]{logP_Kiziltanrho16_3802850078.png}
\includegraphics[width=0.5\textwidth]{logP_Kiziltanrho51_2132034356.png}
\caption{log($P(\vec{t}_k|\text{signal},\rho)$) for different values of $\rho$ for the \FIXME{Kiziltan BNS - change to BBH} source population. The solid black lines denote the mass distribution of the respective source population model.}
\label{fig:P_tkj_BBH}
\end{figure}

\begin{figure}
\includegraphics[width=0.5\textwidth]{lagrangian_fit_Ozel.pdf}
\includegraphics[width=0.5\textwidth]{lagrangian_fit_Kiziltan.pdf}
\caption{Lagrangian fits of $P_{jk}$ vs $\rho$ for a few templates (top: BNS, bottom: \FIXME{Kiziltan BNS - change to BBH}). Dots are the data points, and solid lines are the functions found using a \FIXME{Lagrangian fit - change to cubic interpolation with increased $\rho$ sampling}. The $\rho$ values are found using the Chebychev nodes, with the addition of $\rho=0$.}
\label{fig:Pvsrho}
\end{figure}

@@

>>>>

The direct detections of gravitational waves by the Advanced Laser Interferometer Gravitational-wave Observatory (Advanced LIGO) that were made in late 2015 \cite{Abbott:2016blz, Abbott:2016nmj, TheLIGOScientific:2016pea} mark a monumental breakthrough in physics and astronomy. With additional Advanced LIGO observing runs scheduled and further gravitational-wave observatories expected to come online in in the next several years, the age of gravitational-wave astronomy has truly begun. Advanced LIGO's design sensitivity aims to achieve a factor of three improvement to the signal-to-noise ratio (SNR) for binary black hole systems similar to the detected events \cite{TheLIGOScientific:2016agk}, which are expected to increase the inferred rate of observed compact binary coalescences \cite{Abbott:2016nhf}. With additional detections occurring more and more frequently, we stand to gain a deeper understanding of these tumultuous and dramatic astrophysical events.

@@

In Section \ref{sec:rankingstat}, we begin by describing the ranking statistic that is currently used in one of Advanced LIGO's low-latency searches for compact binary coalescences. The derivation of a ranking statistic that envelops astrophysical source model information is outlined in Section \ref{sec:derivation}, where the various terms are explored in further detail in Sections \ref{sec:P_j} and \ref{sec:P_k}. In Section \ref{sec:P_jk}, we present a discussion of the source-dependent ranking statistic and test it using a BNS and BBH source population model. We conclude in Section \ref{sec:conclusion}.

@@

We consider a gravitational-wave candidate $\vec{s}$; it belongs to an astrophysical source population $\mathcal{S}$ of parameter space dimension $N$. For example, we choose a population that has parameters $\vec\theta$:
   \begin{equation}
   \vec\theta = \{m_1,m_2,\chi_1,\chi_2,\rho\},
   \label{eqn:theta}
   \end{equation}
where $m_i$ and $\chi_i$ are the masses and spins, respectively, of source $i=1,2$ and $\rho$ is the nominal SNR.

@@

The first term in the summation of Eq. \ref{eqn:prob_signal} depends on the astrophysical source population model; therefore, a choice must be made for the source population model. In this work, we will consider two models: a binary neutron star (BNS) population and a binary black hole (BBH) population. The BNS source population we used is from Ozel {\it et al.} \cite{Ozel:2012ax}, where neutron star masses are Gaussian distributed with a centre around 1.33+1.33$M_\odot$ and a dispersion of 0.05$M_\odot$. For the BBH source population, we assume a Gaussian distribution centred around 30+30$M_{\odot}$ with a dispersion of 5$M_\odot$. The population models do not consider the spins of the sources.

The probability density function (PDF) of the source population is given as $P(\vec{\gamma})$, where $\vec{\gamma}=\{\gamma_1,\gamma_2,..., \gamma_N\}$ are the intrinsic parameters of the source population, such as the masses and spins. The probability of that a signal looks like $\vec{t}_j$ (multiplied by some $\rho$) is the following:
	\begin{align}
	P(\vec{t}_j|\text{signal}) &= \int P_j(\vec{\gamma}) \diff\vec{\gamma} \\
	&\approx P_j(\gamma_1) P_j(\gamma_2)...P_j(\gamma_N) \times V_j.
	\end{align}
In the final equality, we assume that the distribution of the source parameters are independent of one another, so the joint probability is the product of the marginals. Because the template bank is made of discrete waveforms, the template bank is turned into a Voronoi diagram, where each template is represented by a Voronoi region. $V_j$ is the $N$-dimensional volume of the $j$th region of the Voronoi diagram, computed using Delaunay triangulation, and it is assumed that $V_j$ is small enough such that $P_j(\gamma_i)$ is constant within the patch. We compute $P(\vec{t}_j|\text{signal})$ for each template in the template bank to find the probability that a system from the given source population will have the same properties as the template. Results for both source populations, BNS and BBH, are shown in Figure \ref{fig:p_j}.

@@

Note that when $\ptjtk=0$, both numerator and denominator equal 0, so Eq. \ref{eqn:p_k} goes to 1, which means that the signal is equally likely to be recovered by any template waveform.

@@

We assume that $P(\rho)\propto\rho^{-4}$. We cannot normalize this because there would be an infinite density of $\rho=0$ sources at infinite distance. Instead, we normalize the above by setting $\rho_{\text{low}}=4$, the trigger threshold for a gravitational-wave search, and $\rho_{\text{high}}=\infty$.

\begin{figure}
\includegraphics[width=0.5\textwidth]{logP_Ozelrho5_18758277176.png}
\includegraphics[width=0.5\textwidth]{logP_Ozelrho11_6127883904.png}
\includegraphics[width=0.5\textwidth]{logP_Ozelrho53_7046503713.png}
\caption{log($P(\vec{t}_k|\text{signal},\rho)$) for different values of $\rho$ for the BNS source population. The solid black lines denote the mass distribution of the respective source population model.}
\label{fig:P_tkj_BNS}
\end{figure}
\begin{figure}
\includegraphics[width=0.5\textwidth]{logP_BBHrho5_18758277176.png}
\includegraphics[width=0.5\textwidth]{logP_BBHrho11_6127883904.png}
\includegraphics[width=0.5\textwidth]{logP_BBHrho53_7046503713.png}
\caption{log($P(\vec{t}_k|\text{signal},\rho)$) for different values of $\rho$ for the BBH source population. The solid black lines denote the mass distribution of the respective source population model.}
\label{fig:P_tkj_BBH}
\end{figure}

\begin{figure}
\includegraphics[width=0.5\textwidth]{lnP_vs_rho_Ozel.pdf}
\includegraphics[width=0.5\textwidth]{lnP_vs_rho_BBH.pdf}
\caption{Lagrangian fits of $P_{jk}$ vs $\rho$ for select templates, which are different for each source population (top: BNS, bottom: BBH). Dots are the data points, and solid lines are the functions fitted with cubic splines. The $\rho$ values are found using the Chebyshev nodes, with the addition of $\rho=0$.}
\label{fig:Pvsrho}
\end{figure}

@@

We are grateful for computational resources provided by the Leonard E Parker
Center for Gravitation, Cosmology and Astrophysics at University of
Wisconsin-Milwaukee.
